---
title: "Sequence Analysis of Neighborhood Racial and Ethnic Changes: The Case of New York City 1980-2020"
author:
  - name: Elizabeth Delmelle
    affiliations:
      - name: University of Pennsylvania
        department: City and Regional Planning
    orcid: 0000-0003-3735-8359
    email: delmelle@upenn.edu
  - name: Eric Delmelle
    affiliations:
      - name: Lehigh University and Vrije Universiteit Brussels
    orcid: 0000-0002-5117-2238
    email: delmelle@gmail.com
date: "2025-08-08"
abstract: |
  This paper demonstrates the application of sequence analysis to develop a typology of racial and ethnic trajectories in New York City neighborhoods from 1980 to 2020 using a reproducible R workflow. Our workflow begins with using an unsupervised classification method, k-means, at each decennial cross-section to derive 6 classes describing the racial and ethnic makeup of neighborhoods during the study period. These classes include four that depict a majority of Black, White, Hispanic, and Asian residents, and two mixed-race classes, Black and Hispanic, and a White majority with a mixture of other races. We then develop a sequence of classes for each census tract over the 5 decennial time stamps. Finally, we derive a longitudinal typology describing the predominant pathways of change using sequence analysis. This resulted in 14 distinct pathways including transitions to Hispanic and Asian majorities emerging from historically White or Black neighborhoods. The findings underscore the gradual nature of neighborhood racial transformations. Our approach is reproducible for researchers wanting to explore and visualize multidimensional neighborhood dynamics.
keywords: [sequence analysis, neighborhood change, New York City]
toc: true
format:
  html:
    code-fold: true
    toc: true
    highlight-style: solarized
    theme: cosmo
    css: styles.css
    code-copy: true
    code-overflow: wrap
execute:
  echo: true
bibliography: "references.bib"
editor_options: 
  markdown: 
    wrap: 72
sidebar: true
---

## Introduction {#sec-intro}

Tracking and understanding neighborhood changes has been a central topic
of urban studies and a fundamental concern for planning practitioners
[@Galster2001; @Landis2016; @ChappleZuk2016]. Neighborhood change can be
comprehended according to various dimensions, from housing stock or
built environment changes to residents' demographic and socioeconomic
composition [@Delmelle2022]. Thus, the study of neighborhood change
involves analyzing multiple attribute dimensions through time for
spatially situated units. Recent scholarship has progressed in the
analytical strategies used to study neighborhood trajectories,
introducing new methods for visualizing and mapping longitudinal
pathways of change for multiple dimensions [@Delmelle2022].

In this article, we demonstrate one such technique, sequence analysis, in an illustrative, reproducible tutorial analyzing neighborhood change. We demonstrate the method using a case study of racial and ethnic changes in New York City census tracts from 1980-2020. While our focus is methodological, the empirical case study highlights the utility of sequence analysis in visualizing, detecting, and exploring longitudinal patterns of neighborhood changes.

Since the 1980s, The United States' demographic profile has become
increasingly diverse in the past several decades, driven by sustained immigration adn by the growing share of births to non-White populations, influencing demographic shifts in the overall population structure [@frey2022new]. Analyses following the release of the 2010 decennial
census showed that increasing national diversity resulted in differing
neighborhood trajectories, largely contingent on the broader
metropolitan context [@Terbeck2023; @Wright2014].

Much of the empirical scholarship on neighborhood racial and ethnic
changes has developed indices to categorize the makeup or diversity of
racial and ethnic groups and then explored changes in a neighborhood's
diversity categorization over time. For example, a neighborhood might
transition from 'low diverse' to 'moderately diverse' from one decade to
the next [@farrell2011racial; @Wright2014]. Alternately, another set of
techniques aims to describe the longitudinal sequences or trajectories
that the racial and ethnic groups in a neighborhood have followed. Three
methods have been adopted in the literature for this purpose: statistical curve fitting, time-series clustering, and sequence analysis. With curve-fitting models like growth mixture models or latent growth models, mathematical functions fit each racial and
ethnic group under study that summarize the predominant trends
[@Zwiers2018; @HippKim2023]. Time series clustering is an unsupervised classification technique aimed at clustering continuous longitudinal data [@delmelle2025time]. Finally,in sequence analysis, neighborhoods are first grouped into similar categorical clusters at each time stamp. 
Then, each a sequence of neighborhood categories is constructed over time. Finally, the sequences
of these clusters are grouped using a sequence alignment technique
[@Delmelle2016; @GonzalezLeonardo2023]. 

While all three approaches - curve fitting, time-series clustering, and sequence analysis aim to characterize change over time, they differ in what types of data and research questions they are best suited for. Curve fitting methods are well-suited for summarizing trends in continuous outcomes, especially when the trajectory is expected to follow a smooth or parametric form, which must be specified *a priori*. Time-series clustering also retains the continuous nature of longitudinal data, but must be performed on single variables at a time. Cross-sectional classes of trajectories can be constructed to form multivariate typologies [@delmelle2025time], however, as the number of variables increases, this becomes an arduous workflow. 

Sequence analysis is oriented towards categorical trajectories that can be constructed using multiple input variables, as demonstrated in this article. The method is particularly useful in visualizing and analyzing the timing, order, and transitions between qualitatively distinct neighborhood states rather than in the smoothness of changes.

In this article, we explore trajectories of neighborhood racial and
ethnic changes in the largest and one of the most diverse cities in the
United States, New York City, using longitudinal census data up to the
latest 2020 decennial release. This notebook showcases a workflow that
introduces sequence analysis to the study of multidimensional
neighborhood changes. We begin by processing the raw longitudinal census
data, then performing a k-means classification, and finally classify and
map sequences clusters. Our case study illustrates the gradual
progression that neighborhoods follow in when undergoing racial and
ethnic transformations. We observe an overall decline in the share of
neighborhoods categorized by a large White majority population, in
exchange for increasing diversity, especially Hispanic and Asian
populations.

## Computational environment {#sec-compEnv}

The main libraries used in this paper include `dplyr` for processing
tabular census data, `sf` for mapping the resulting clusters, `cluster`
for performing the `k-means` cluster analysis. Finally, to perform the
sequence analysis, we use `TraMineR` [@gabadinho2011analyzing]. This is
a popular and continually updated R package for performing sequence
analysis for a host of social science applications, including analyses
of neighborhood change [@Delmelle2016; @delmelle2017differentiating;
@patias2020scalable].

```{r loadlibraries, include = FALSE, warning=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# For data management
if (!require('knitr')) install.packages('knitr'); library('knitr')
if (!require('rmarkdown')) install.packages('rmarkdown'); library('rmarkdown')
if (!require('dplyr')) install.packages('dplyr'); library('dplyr')

#for reading in data
if (!require('here')) install.packages('here'); library('here')

#for reading in census data
if (!require('tidycensus')) install.packages('tidycensus'); library('tidycensus')
if (!require('tigris')) install.packages('tigris'); library('tigris')
options(tigris_class = "sf") # returns data in sf format

#for data table formatting
if (!require('knitr')) install.packages('knitr'); library('knitr')

#for data table pivoting
if (!require('tidyverse')) install.packages('tidyverse'); library('tidyverse')


#mapping packages
if (!require('sp')) install.packages('sp'); library('sp')
if (!require('sf')) install.packages('sf'); library('sf')

#For data visualization
if (!require('ggplot2')) install.packages('ggplot2'); library('ggplot2')

#for facet plots
if (!require('gridExtra')) install.packages('gridExtra'); library('gridExtra')

#For sequence clustering
if (!require('TraMineR')) install.packages('TraMineR'); library('TraMineR')

# For k-means and hierarchical cluster analysis
if (!require('cluster')) install.packages('cluster'); library('cluster')

#For visualizing k-means outputs
if (!require('factoextra')) install.packages('factoextra'); library('factoextra')

# For creating heat map to describe k-means cluster results
if (!require('pheatmap')) install.packages('pheatmap'); library('pheatmap')

# For creating heat map to describe k-means cluster results
if (!require('RColorBrewer')) install.packages('RColorBrewer'); library('RColorBrewer')

# Other packages (cowplot:inset maps; extrafont for additional sans serif fonts)
if (!require('cowplot')) install.packages('cowplot'); library('cowplot')
if (!require('patchwork')) install.packages('patchwork'); library('patchwork')
if (!require('extrafont')) install.packages('extrafont'); library('extrafont')
loadfonts(device = "win") 
if (!require('ggspatial')) install.packages('ggspatial'); library('ggspatial')
loadfonts(device = "win") 
```

## Data {#sec-data}

We use decennial census tract data to examine neighborhood racial and
ethnic changes. Census tracts serve as imperfect, yet well-used
neighborhood proxies. Census tract boundaries change over time, further
complicating the study of population dynamics within these boundaries.
There are several sources of data that have been harmonized using
interpolation techniques to a consistent set of boundaries over time. We
use the Longitudinal Tract Database (LTDB) which uses areal and
population interpolation techniques alongside ancillary data on water
cover to derive estimates [@logan2014interpolating]. Analyses of the
errors produced by three popular longitudinal data providers suggest
that LTDB performs similarly to the dataset produced by the National
Historic Geographic Information System (NHGIS) and both perform better
than the Neighborhood Change Database which relies solely on areal
interpolation without the inclusion of ancillary data
[@logan2014interpolating]. Therefore, for this type of analysis either
the LTDB or NHGIS would be suitable dataset for this analysis.

We obtained the full count decennial data from LTDB from 1980-2020 from
the [Diversity and Disparties project at Brown
University](https://s4.ad.brown.edu/projects/diversity/Researcher/LTBDDload/DataList.aspx).
The census variables have been interpolated to 2010 tract boundaries.
Because the coding of census race and ethnicity changes over time, we
opted to begin in 1980 as 1970, the earliest dataset available, did not
record a count of Latino or Hispanic residents. The raw data contains
all census tracts throughout the United States. For visualization
purposes, we also import a shapefile of 2010 census tract boundaries
using the `tidycensus` package and setting the geometry to `true`.

```{r importdata, warning=FALSE, cache = TRUE, message=FALSE}
setwd(here())  #current working directory

#csv tables for longitudinal data
census20<- read.csv("data/ltdb_std_2020_fullcount.csv")
census10<- read.csv("data/LTDB_Std_2010_fullcount.csv")
census00<- read.csv("data/LTDB_Std_2000_fullcount.csv")
census90<- read.csv("data/LTDB_Std_1990_fullcount.csv")
census80<- read.csv("data/LTDB_Std_1980_fullcount.csv")

#geometry data
#filter for NYC counties (5 boroughs)
nyc_counties <- c("Kings", "Queens", "New York", "Richmond", "Bronx")
tract <- get_decennial(geography = "tract",
                       variables = "P001001",
                       year = 2010,
                       state = "NY",
                       county = nyc_counties,
                       geometry = TRUE,
                       progress = FALSE)
```

We next calculate the share of `White`, `Black`, `Hispanic`, and `Asian`
residents in each tract for each decade from the raw count using the
total population as the denominator. We filter out tracts where the
population is `0` and select only the relevant columns to create our
data frame. We then join all columns from the five decennial data frames
into one data frame called `census_all` and finally select only census
tracts from the five counties that comprise New York City's five
boroughs: Bronx County, Kings County (Brooklyn), New York County
(Manhattan), Queens County, Richmond County (Staten Island).

```{r calculatepercents, warning=FALSE, message=FALSE}
census80 <- census80 %>% filter (POP80 >0)
census80$perwhite80 <- census80$NHWHT80/census80$POP80
census80$perblack80 <- census80$NHBLK80/census80$POP80
census80$perhisp80 <- census80$HISP80/census80$POP80
census80$perasian80 <- census80$ASIAN80/census80$POP80
census80<- census80 %>% select(c("TRTID10","perwhite80", "perblack80", "perhisp80", "perasian80"))

census90 <- census90 %>% filter (POP90 >0)
census90$perwhite90 <- census90$NHWHT90/census90$POP90
census90$perblack90 <- census90$NHBLK90/census90$POP90
census90$perhisp90 <- census90$HISP90/census90$POP90
census90$perasian90 <- census90$ASIAN90/census90$POP90
census90<- census90 %>% select(c("TRTID10","state","county","perwhite90", "perblack90", "perhisp90", "perasian90"))

census00 <- census00 %>% filter (POP00 >0)
census00$perwhite00 <- census00$NHWHT00/census00$POP00
census00$perblack00 <- census00$NHBLK00/census00$POP00
census00$perhisp00 <- census00$HISP00/census00$POP00
census00$perasian00 <- census00$ASIAN00/census00$POP00
census00<- census00 %>% select(c("TRTID10","perwhite00", "perblack00", "perhisp00", "perasian00"))

census10 <- census10 %>% rename("TRTID10" = "tractid")
census10$perwhite10 <- census10$nhwht10/census10$pop10
census10$perblack10 <- census10$nhblk10/census10$pop10
census10$perhisp10 <- census10$hisp10/census10$pop10
census10$perasian10 <- census10$asian10/census10$pop10
census10<- census10 %>% select(c("TRTID10","perwhite10", "perblack10", "perhisp10", "perasian10"))

census20 <- census20 %>% rename("TRTID10" = "TRTID2010")
census20$perwhite20 <- census20$nhwt20/census20$pop20
census20$perblack20 <- census20$nhblk20/census20$pop20
census20$perhisp20 <- census20$hisp20/census20$pop20
census20$perasian20 <- census20$asian20/census20$pop20
census20<- census20 %>% select(c("TRTID10","perwhite20", "perblack20", "perhisp20", "perasian20"))

#join all data frames from each decade
census_all<- census90 %>% left_join(census00) %>% left_join(., census10) %>% left_join(., census20)%>% left_join(., census80)

#Select NYC Counties. These include Bronx County, Kings County (Brooklyn), New York County (Manhattan), Queens County, Richmond County (Staten Island)

census_select <- census_all %>% filter((state == "NY" & county == "Bronx County")|
                              (state == "NY" & county == "Kings County")|
                              (state == "NY" & county == "New York County")|
                              (state == "NY" & county == "Queens County")|
                              (state == "NY" & county == "Richmond County"))

##remove NA values and state and county columns
census_nyc <- na.omit(census_select)%>% select(-state, -county)

#Select only the tractID column from the shapefile. Rename the field for ease of joining and convert to double to match the csv data.
tract<- tract %>% select("GEOID")
tract<- rename(tract, TRTID10 = GEOID)
tract$TRTID10<- as.double(tract$TRTID10)
```

```{r extract_county_boundaries, warning=FALSE, cache = TRUE, message=FALSE}
# Set tigris options to return the data in sf format (spatial data frame)
options(tigris_class = "sf")

# Download county boundaries for New York state
ny_counties <- counties(state = "NY", cb = TRUE,
                         progress = FALSE)

# Filter for only the New York City counties
nyc_counties <- ny_counties %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

# View the structure of the data
print(nyc_counties)
```

## Basic conceptual intuition {#sec-intuition}

### Categorizing Neighborhoods Using *k*-means

Sequence analysis requires categorical input states. Given that neighborhood demographic and socioeconomic data is very often continuous, the first step is to transform the data into discrete categories for each year of the analysis. In this example, we opt to use a data-driven, unsupervised classification approach for this stage by applying *k*-means to cluster the racial and ethnic composition of each census tract in each decade into one of several mutually exclusive categories. This step serves as a preprocessing stage, not an analysis of changes itself. 

As an alternative, for the case of racial and ethnic compositions, for example, pre-specified thresholds (e.g., >50% of a population group) could be used to define categories like "majority Black" or "majority Hispanic". This approach works well in situations where clear theoretical cutoffs exist. However, other applications involving socio-economic indicators or multivariate neighborhood characteristics may not lend themselves to clear analyst-defined cutoffs, making a data-driven approach preferable. Since the purpose of this article is illustrative, we proceed with an overview of implementing *k*-means to construct the initial discrete classes of neighborhoods from the original racial and ethnic makeup data. However, if the analyst already has theoretically grounded thresholds, they can bypass this step and proceed directly to the sequence analysis.

In the context of neighborhood racial and ethnic classification, [@reibel2011neighborhood] introduced the idea of using an unsupervised
classification approach for studying neighborhood racial change as an
alternative to the use of neighborhood diversity indices. The objective
of the *k*-means algorithm is to group observations in such a way that
maximizes the similarity of observations within groups or clusters while
maximizing the dissimilarity between each cluster. In other words, the
goal is to group neighborhoods so that those most similar to each other
according to their racial and ethnic makeup are assigned to the same
cluster and the clusters themselves are distinct from one another in
terms of their makeup. 

With the *k*-means algorithm, the number of clusters, *k* must be determined *a priori*. To make this determination,
it is customary to evaluate multiple solutions using various fit
statistics in conjunction with domain and geographic knowledge of the
data [@delmelle2015five]. It is also commonly recommended that input
variables first be normalized to avoid placing unequal emphasis on
variables that may be on different measurement scales. However, in our
case study, all of the racial and ethnic variables represent percentages
of the population and so this step is not performed in the case study.

Our ultimate goal is to understand the major pathways of neighborhood
change and so each neighborhood will be classified five times for 1980,
1990, 2000, 2010, and 2020 to establish its longitudinal sequence. To
ensure that the clusters are temporally stable, we will perform the
clustering for all years at once. New neighborhood typologies may emerge
over time with this approach. In that instance, only tracts from the
later years would be assigned to the new cluster.

### Identifying pathways of change using sequence analysis

Like *k*-means, sequence analysis is an unsupervised classification
technique but instead of grouping observations based on cross-sectional similarity, it clusters them based on the similarity of longitudinal categorical sequences. A central component of sequence analysis is defining how similar or dissimilar two sequences are. There are multiple methods to compute sequence dissimilarity. [@studer2016matters] provides a comprehensive overview of various techniques for determining sequences dissimilarity for social science
applications. 

In this case study, our goal is to group neighborhoods that follow similar racial and ethnic change trajectories, specifically in which order of categories traverse through. This gives us an overview of the various pathways of change neighborhoods may take. For example sequences that progress from a majority White composition to Mixed Race and eventually to majority Hispanic may represent one frequent change pathway. Because our sequences are measured at fixed intervals (every decade from 1980 to 2020), and each neighborhood has the same number of observations, we are less concerned with irregular timing or duration, which are often central considerations in life-course studies. 

To prioritize this kind of ordered similarity, we use OMstrans, a variation of the popular Optimal Matching (OM) algorithm [@delmelle2017differentiating]. OM treats sequences as strings and calculates the 'edit distance', or minimal cost to transform one sequence into another using insertions, deletions, and substitutions. In OMstrans, the algorithm operations not on the raw states, but on the transitions between states by merging eaech state in a sequence with its predecessor. This approach emphasizes how one state leads into another and helps preserve the sequencing of change.  

For example, one neighborhood may have a
sequence of *White*, *White*, *Mixed Race*, *Hispanic* while another
neighborhood might follow the sequence *White*, *White*, *Mixed Race*,
*Mixed Race*. Between these two strings, there is one entry that
differs: the final state. To transform one to another, we could
substitute *Hispanic* for *Mixed Race* in the second sequence, and the
dissimilarity of the two sequences would be equal to the cost of that
substitution. 

In the OMstrans variant, distances between sequences of
transitions are computed. This means that each state is merged with its
previous state to create a subsequence. In the previous illustrative
example, the first sequence becomes (*White-White*, *White-MixedRace*,
*MixedRace-Hispanic*) and the OM cost evaluation is then applied to
these subsequences. 

The OMstrans approach allows us to incorporate empirical transition probabilities into the substitution cost matrix. In this way, frequent transitions (e.g., *White* to *Mixed*) are penalized less than rare ones. Insertions and deletions can be discoraged by assigning a higher cost than the empirical transitions so that disruptions to the temporal alignment of sequences is discouraged. The balance between preserving state similarity and order is further governed by the parameter *w*, the origin-transition parameter. When *w* = 1, OMstrans approximates the
traditional OM algorithm. A lower value places greater emphasis on the
ordering or sequencing of events than on the specific states themselves.

## Application {#sec-application}

### Study Area

Our study area consists of census tracts within the five Boroughs of New
York (see Figure below for an overview map).

```{r visualize-area, warning=FALSE, cache=TRUE, message=FALSE, fig.cap="Overview map of New York City five boroughs and surrounding counties", fig.width=10, fig.height=8}

# Filter for only the New York City counties
nyc_counties <- ny_counties %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond")) 

# Modify labels for specific counties
nyc_counties$label_main <- ifelse(nyc_counties$NAME == "New York", "New York", 
                                  ifelse(nyc_counties$NAME == "Kings", "Kings",
                                         ifelse(nyc_counties$NAME == "Richmond", "Richmond", 
                                                nyc_counties$NAME)))

nyc_counties$label_paren <- ifelse(nyc_counties$NAME == "New York", "(Manhattan)", 
                                   ifelse(nyc_counties$NAME == "Kings", "(Brooklyn)",
                                          ifelse(nyc_counties$NAME == "Richmond", "(Staten Island)", 
                                                 "")))

# Calculate centroids for each county to get coordinates for labels
centroids <- st_centroid(nyc_counties)
coords <- st_coordinates(centroids)

# Add the coordinates to the nyc_counties data frame
nyc_counties <- nyc_counties %>%
  mutate(X = coords[, 1],
         Y = coords[, 2])

# Adjust coordinates manually for specific counties
nyc_counties <- nyc_counties %>% 
  mutate(
    X = ifelse(NAME == "New York", X - 0.11, 
               ifelse(NAME == "Richmond", X + 0.012, X)), # Move "New York" left, Richmond right
    Y = ifelse(NAME == "New York", Y + 0.005, 
               ifelse(NAME == "Richmond", Y + 0.03, Y)) # Adjust Richmond label higher
  )

#you can mask water area if you would like.
#nyc_counties <- nyc_counties %>%  erase_water(area_threshold = 0.9)

zoomed_map <- ggplot() +
  geom_sf(data = ny_counties, fill = "gray85", color = "white", size = 1) +
  geom_sf(data = nyc_counties, fill = "#85B0A9", color = "white", size = 3) +
  geom_segment(aes(x = -73.98, y = 40.775, xend = -74.03, yend = 40.775), 
               color = "gray25", size = 0.5) +
  geom_text(data = nyc_counties, aes(x = X, y = Y, label = label_main), 
            color = "black", size = 4, family = "Arial", fontface = "bold") +
  geom_text(data = nyc_counties, aes(x = X, y = Y - 0.02, label = label_paren), 
            color = "gray25", size = 3.75, family = "Arial") +
  geom_text(aes(x = -73.6, y = 40.725, label = "Nassau"), 
            color = "gray55", size = 4, family = "Arial", fontface = "bold") +
  geom_text(aes(x = -73.8, y = 40.98, label = "Westchester"), 
            color = "gray55", size = 4, family = "Arial", fontface = "bold") +
  geom_text(aes(x = -73.6, y = 40.705, label = "(Long Island)"), 
            color = "gray70", size = 3.5, family = "Arial", fontface = "bold") +
  annotation_scale(location = "bl", width_hint = 0.18, 
                   text_cex = .9,
                   bar_cols = c("gray75", "white"),
                   style = "bar",
                   unit_label = "km") +
  coord_sf(xlim = c(-74.25, -73.5), ylim = c(40.4, 41.05), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

# Create the large map for NY and surrounding counties with text annotation
inset_map <- ggplot() +
  geom_sf(data = ny_counties, fill = "gray90", color = NA, size = 1) + # All counties in gray, including Nassau and Westchester
  geom_sf(data = nyc_counties, fill = "#569289", color = "#569289", size = 5) + # NYC counties in darker teal
  annotate("text", x = -76.2, y = 42.8, label = "New York State", 
           size = 5.25, color = "gray49", angle = 0, 
           alpha = 0.7, family = "Arial") + # Add text label
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    plot.title = element_text(hjust = 0), # Align title to the left
    plot.subtitle = element_text(hjust = 0), # Align subtitle to the left
    axis.title = element_blank(), # Remove axis titles
    axis.text = element_blank(), # Remove axis text (labels)
    plot.margin = margin(0, 0, 0, 0) # Tighten margins
  )

# Combine the maps, placing the inset map within the zoomed-in map
final_plot <- ggdraw() +
  draw_plot(zoomed_map) +
  draw_plot(inset_map, x = 0.07, y = 0.65, width = 0.3, height = 0.3) # Adjust position and size of the inset

# Display the combined plot
print(final_plot)
```

To begin our case study, we start by preparing the data for the
*k*-means clustering. This involves pivoting the data frame so that each
census tract is represented with five distinct rows, once for each
decennial value. To do so, we pivot from a wide to a long format and
select out just the four race and ethnicity values to be used in the
clustering.

```{r kmeans, warning=FALSE, cache=TRUE, message=FALSE}
# Convert the data frame from wide to long format
census_long <- census_nyc %>%
  pivot_longer(cols = starts_with("per"), 
               names_to = c(".value", "year"), 
               names_pattern = "per(\\w+)(\\d{2})") %>%
  mutate(year = case_when(
    year == "80" ~ 1980,
    year == "90" ~ 1990,
    year == "00" ~ 2000,
    year == "10" ~ 2010,
    year == "20" ~ 2020,
    TRUE ~ as.integer(year)
  ))


data_for_clustering <- census_long %>%
  select(white, black, hisp, asian)
```

As explained above, the *k*-means clustering procedure requires that the
number of clusters, *k* be specified *a priori*. There are fit
statistics that attempt to determine the optimal number of clusters,
considering the similarity of observations within each cluster and the
distinctiveness of the clusters from each other. However, these
mathematically-derived solutions are devoid of any contextual or
theoretical understanding of the problem under study. Therefore, the
selection of *k* often becomes more akin to art than a science
[@von2012clustering], considering the objective of the study. We begin
with two common data-driven approaches that explore multiple clustering
solutions for different *k* values and then examines the *within sum of
squares* (WSS) and the *average silhouette score* for each solution. The
WSS assesses how compact a clustering solution is, or how homogeneous
the observations assigned to each cluster are, and the average
silhouette score measures how well separated each cluster is from each
other. Our objective is to derive a typology of neighborhoods according
to their racial and ethnic makeup, for four groups: percent White,
Black, Hispanic, and Asian. The Figure below shows the results of the
clustering analysis; the Elbow method plot suggests the optimal number
of clusters, while the Silhouette method plot helps validate the cluster
separation.

```{r kmeans1, warning=FALSE, cache=TRUE, message=FALSE}
# Now do the k-means clustering on all

data_for_clustering <- census_long %>%
  select(white, black, hisp, asian)

# Function to calculate total within-cluster sum of squares for different k
wss <- function(k) {
  kmeans(data_for_clustering, k, nstart = 10)$tot.withinss
}

# Compute and plot wss for k = 1 to k = 10
k.values <- 1:10
wss_values <- map_dbl(k.values, wss)

# Elbow method plot
plot(k.values, wss_values, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares")

# Silhouette method for determining the optimal number of clusters
fviz_nbclust(data_for_clustering, kmeans, method = "silhouette")
```

### Exploring *k*-means Cluster Solutions

According to these plots, the mathematically optimal number of
neighborhood clusters for racial makeup is three. We can further explore
the makeup of neighborhoods within these three clusters a few ways to
determine if, in fact, three clusters provides a meaningful segmentation
of four distinct racial and ethnic groups. In the plots below, we can
visualize the average silhouette value for each cluster. The
*Silhouette* values range from -1 to 1; values close to 1 suggest that
the observations are well clustered while negative values suggest that
an observation might be assigned to the wrong cluster. From the plot,
Clusters 2 and 3 appears to be the most cohesive clusters, with average
silhouette widths of 0.64, while cluster 1 has some potentially poorly
classified neighborhoods. Descriptions of the racial and ethnic makeup
of the clusters are obtained from the associated stacked bar charts. We
can see that Cluster 1 is characterized as being nearly 50 percent
Hispanic, with near equal shares of Whites, Blacks, and Asians. Cluster
2 is majority Black with approximately 15 percent Hispanics and few
Whites and Asians. Finally, Cluster 3 is majority White. Therefore, this
segmentation provides us with clusters indicating the dominant racial
groups, but may miss some nuances of other racial and ethnic
neighborhood compositions. We can therefore explore how increasing *k*
may portray a richer portrait of neighborhood demographic profiles. The
Figure below illustrates the Silhouette Analysis to assess cluster
separation, and the demographic makeup of the three clusters (bottom
Figure).

```{r threeclusters, warning=FALSE, cache=TRUE, message=FALSE, fig.cap="Cluster Analysis Results: Silhouette Analysis and Cluster Demographic Makeup"}
# Assume the optimal number of clusters (k) is 3 from the previous steps
set.seed(123)
kmeans_result <- kmeans(data_for_clustering, centers = 3, nstart = 25)

# Add the cluster assignments to the original data
census_long$cluster <- kmeans_result$cluster

# Silhouette Analysis
sil <- silhouette(kmeans_result$cluster, dist(data_for_clustering))
fviz_silhouette(sil)

# Extract silhouette information to a data frame (df)
sil_df <- as.data.frame(sil[, 1:3])
colnames(sil_df) <- c("Cluster", "Silhouette Width", "Neighboring Cluster")

# Cluster profiles
cluster_profiles <- census_long %>%
  group_by(cluster) %>%
  summarise(across(c(white, black, hisp, asian), ~ round(mean(.), 2)))

# Print the table as a formatted table
kable(cluster_profiles, caption = "Cluster Profiles: Average Demographics", 
      col.names = c("Cluster", "White", "Black", "Hispanic", "Asian"),
      format = "markdown")

# Reshape the data from wide to long format
cluster_profiles_long <- cluster_profiles %>%
  pivot_longer(cols = c("white", "black", "hisp", "asian"), 
               names_to = "Demographic", values_to = "Proportion")


# Create the stacked bar chart
ggplot(cluster_profiles_long, aes(x = factor(cluster), y = Proportion, fill = Demographic)) +
  geom_bar(stat = "identity") +
  labs(title = "Cluster Demographic Makeup - 3 clusters", x = "Cluster", y = "Proportion") +
  scale_fill_manual(
  values = c(
    "white" = "#F0E442",     # yellow
    "black" = "#0072B2",     # blue
    "hisp"  = "#D55E00",     # orange/red
    "asian" = "#CC79A7"      # magenta
  ),
  labels = c(
    "white" = "White",
    "black" = "Black",
    "hisp" = "Hispanic",
    "asian" = "Asian"
  ),
  name = "Demographic Group"
) +  # Adjust color palette if desired
  theme_minimal()
```

Next, we compare 4, 5, and 6 cluster solutions. We can see that the
average silhouette of the solutions declines as the number of clusters
increases. The demographic profiles show several new neighborhood
typologies emerge with more clusters added. With a four cluster
solution, we observe neighborhood typologies for each of the three
dominant racial and ethic groups: White, Hispanic, and Black along with
one mixed neighborhood type, shown in cluster 4. As expected, that
cluster displays the lowest silhouette value, with some potentially
mis-classified neighborhoods. The 5 cluster solution adds a cluster
showing a majority Asian population, alongside two majority White
populations - one showing more diversity than the other), and a majority
Black, and Hispanic group. Finally, a 6 cluster solution shows more
racially mixed groups, but at the expense of less well-defined or
separated clusters. In the code below, each variable name ends with a
number (4, 5 or 6), indicating which number of clusters (k) it
represents. The results displays the silhouette analysis results for
clustering into (a) four clusters, (b) five clusters, and (c) six
clusters, illustrating the cohesion and separation of clusters at
different sizes.

```{r silhouette-plots, warning=FALSE, cache=TRUE, message=FALSE,  fig.cap="Silhouette Analysis for Different Cluster Sizes: (top) Four Clusters, (middle) Five Clusters, and (bottom) Six Clusters."}
# Perform k-means clustering for 4, 5, and 6 clusters
set.seed(123)
kmeans_4 <- kmeans(data_for_clustering, centers = 4, nstart = 25)
kmeans_5 <- kmeans(data_for_clustering, centers = 5, nstart = 25)
kmeans_6 <- kmeans(data_for_clustering, centers = 6, nstart = 25)

# Add cluster assignment to original data
census_long$cluster_4 <- kmeans_4$cluster
census_long$cluster_5 <- kmeans_5$cluster
census_long$cluster_6 <- kmeans_6$cluster

# Silhouette analysis for 4, 5, and 6 clusters
sil_4 <- silhouette(kmeans_4$cluster, dist(data_for_clustering))
sil_5 <- silhouette(kmeans_5$cluster, dist(data_for_clustering))
sil_6 <- silhouette(kmeans_6$cluster, dist(data_for_clustering))

# Plot silhouette for 4 clusters
plot_4 <- fviz_silhouette(sil_4) + ggtitle("Silhouette Plot - Four Clusters")

# Plot silhouette for 5 clusters
plot_5 <- fviz_silhouette(sil_5) + ggtitle("Silhouette Plot - Five Clusters")

# Plot silhouette for 6 clusters
plot_6 <- fviz_silhouette(sil_6) + ggtitle("Silhouette Plot - Six Clusters")

# Combine the three plots into a faceted view
grid.arrange(plot_4, plot_5, plot_6, ncol = 1)
```

The figure illustrates the demographic composition of clusters for
different clustering solutions: (a) four clusters, (b) five clusters,
and (c) six clusters, highlighting how demographic groups are
distributed across various cluster labels.

```{r explore-demographics, warning=FALSE, cache=TRUE, message=FALSE, fig.cap="Geographic variation of Labeled Clusters for Different Clustering Solutions."}
# We need to hard code the labels to be able to compare the demographic profile of each solution. This is because R randomly assigns the label each time, even if the clustering solution is the same because we set the seed.
label_clusters <- function(cluster_profiles) {
  cluster_profiles %>%
    mutate(
      label = case_when(
        white > 0.75 ~ "3 White",
        hisp > 0.45 ~ "2 Hispanic",
        black > 0.75 ~ "1 Black",
        asian > 0.45 ~ "5 Asian",
        white < 0.56 & hisp > 0.20 & black < 0.20 ~ "4 Mixed",
        black > 0.49 & hisp > 0.25 & white < 0.15 ~ "6 Black & Hispanic",
        TRUE ~ "Other"  # Default label if none of the conditions are met
      )
    )
}

# Create cluster profiles and label them according to your custom rules
cluster_profiles_4 <- census_long %>%
  group_by(cluster_4) %>%
  summarise(across(c(white, black, hisp, asian), ~ round(mean(.), 2))) %>%
  mutate(`Clustering_Solution` = "4 Clusters", Cluster = cluster_4) %>%
  label_clusters()

cluster_profiles_5 <- census_long %>%
  group_by(cluster_5) %>%
  summarise(across(c(white, black, hisp, asian), ~ round(mean(.), 2))) %>%
  mutate(`Clustering_Solution` = "5 Clusters", Cluster = cluster_5) %>%
  label_clusters()

cluster_profiles_6 <- census_long %>%
  group_by(cluster_6) %>%
  summarise(across(c(white, black, hisp, asian), ~ round(mean(.), 2))) %>%
  mutate(`Clustering_Solution` = "6 Clusters", Cluster = cluster_6) %>%
  label_clusters()

# Combine all cluster profiles into one table
cluster_profiles_combined <- bind_rows(cluster_profiles_4, cluster_profiles_5, cluster_profiles_6)

# Map the labels back to the original tracts by joining based on cluster assignments
census_long <- census_long %>%
  left_join(cluster_profiles_4 %>% select(Cluster, label) %>% rename(label_4 = label), by = c("cluster_4" = "Cluster")) %>%
  left_join(cluster_profiles_5 %>% select(Cluster, label) %>% rename(label_5 = label), by = c("cluster_5" = "Cluster")) %>%
  left_join(cluster_profiles_6 %>% select(Cluster, label) %>% rename(label_6 = label), by = c("cluster_6" = "Cluster"))

# Reshape data for plotting
cluster_profiles_long <- cluster_profiles_combined %>%
  pivot_longer(cols = c("white", "black", "hisp", "asian"),
               names_to = "Demographic", values_to = "Proportion")

# Plot stacked bar charts of demographic makeup for each labeled cluster
ggplot(cluster_profiles_long, aes(x = label, y = Proportion, fill = Demographic)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(Clustering_Solution ~ ., scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c(
      "white" = "#F0E442",     # yellow
      "black" = "#0072B2",     # blue
      "hisp"  = "#D55E00",     # red/orange
      "asian" = "#CC79A7"      # magenta
    ),
    labels = c(
      "white" = "White",
      "black" = "Black",
      "hisp" = "Hispanic",
      "asian" = "Asian"
    ),
    name = "Demographic Group"
  ) +
  labs(
    title = "Demographic Makeup of Labeled Clusters (4, 5, and 6 Clusters)",
    x = "Cluster Label",
    y = "Proportion of Demographic Group"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 13, face = "bold"),
    strip.text.y = element_text(angle = 0)
  )
```

We can explore how this plays out for a specific observation. For
example, take the first census tract in the data frame for the year
`2020`. This tract's racial composition was `58%` Black and `30%`
Hispanic with small shares of Whites and Asians. With the four cluster
solution, this tract was classified into class 1, `Majority Black`. For
the five cluster solution, it was also classified as `Majority Black`,
and for the six cluster solution, `Black and Hispanic`. In 1990, that
same tract was `34%` black and `61%` Hispanic, resulting in the tract
being classified as `Majority Hispanic` by all three clustering
solutions. When looking at change over time, the neighborhood will
either be registered as transitioning from `Majority Hispanic` to either
`Majority Black` or `Black and Hispanic`, depending on the final cluster
solution. In this instance, the 6 cluster solution provides a more
accurate portrayal of the dynamics - while the neighborhood did
technically become majority Black, there is still a significant Hispanic
presence, a detail that would have been omitted by limiting the number
of groups.

Finally, we can examine the spatial distribution of these clustering
solutions to help aid in the final determination for the sequence
analysis. Since we did the cluster analysis on all decades, for this
purpose, we will contrast the 4, 5, and 6 cluster solutions just for
2020. All three maps depict a similar spatial pattern, but with a more
fragmented pattern in the case of the 6 cluster solution. For example,
the first two maps show contiguous tracts of the predominantly Black
cluster, but the third map shows the emergence of the
`Black and Hispanic` group largely forming on the outskirts of this
spatial cluster. Another aparent distinction is the large spatial
cluster of the `Asian` cluster in Queens, which had been labeled as
`Mixed` in the 4 cluster solution.

To better understand details on racial neighborhood transitions, we will
go with the larger number of clusters, 6, despite the mathematical
preference for a 3 cluster solution. Our result illustrates the
demographic distribution of labeled clusters for different clustering
solutions: four clusters, five clusters, and six clusters.

```{r demographic-distribution, warning=FALSE, cache=TRUE, message=FALSE, fig.cap="Demographic Distribution of Labeled Clusters for Different Clustering Solutions: Four Clusters (top), Five Clusters (middle), and Six Clusters (bottom)."}
# Filter data for the year 2020 and merge labels with tract data for mapping
tract_clusters_2020 <- tract %>% erase_water(area_threshold = 0.9) %>%
  left_join(census_long %>%
              filter(year == 2020) %>%  # Filter for the year 2020
              select(TRTID10, cluster_4, cluster_5, cluster_6, label_4, label_5, label_6), by = "TRTID10") %>%
  pivot_longer(cols = c("cluster_4", "cluster_5", "cluster_6"),
               names_to = "ClusterSolution", values_to = "Cluster") %>%
  pivot_longer(cols = c("label_4", "label_5", "label_6"),
               names_to = "LabelSolution", values_to = "Label") %>%
  filter(str_replace(ClusterSolution, "cluster_", "") == str_replace(LabelSolution, "label_", "")) %>%
  mutate(Label = factor(Label)) %>%
  mutate(ClusterSolution = recode(ClusterSolution,
                                  "cluster_4" = "4 Clusters",
                                  "cluster_5" = "5 Clusters",
                                  "cluster_6" = "6 Clusters")) %>%
  mutate(ClusterSolution = factor(ClusterSolution, levels = c("4 Clusters", "5 Clusters", "6 Clusters")))



base_plot <- ggplot(tract_clusters_2020) +
  geom_sf(aes(fill = Label), color = NA) +
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +
  facet_wrap(~ ClusterSolution, nrow = 1, strip.position = "bottom") +
  scale_fill_manual(
    values = c("3 White" = "#F0E442", "2 Hispanic" = "#D55E00", 
               "1 Black" = "#0072B2", "5 Asian" = "#CC79A7", 
               "4 Mixed" = "#009E73", "6 Black and Hispanic" = "#56B4E9"),
    labels = c("Black", "Hispanic", "White", "Asian", "Mixed", "Black & Hispanic")
  ) +
  labs(
    title = "Racial Cluster Maps by Number of Groups",
    x = "", y = "", fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text = element_text(size = 12, face = "italic", color = "gray20"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.width = unit(1.2, "cm"),
    legend.key.height = unit(0.4, "cm")
  ) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE, label.position = "bottom"))

grid.arrange(base_plot)
```

We can begin with a simple exploration of the spatial changes over time
in the maps showing the six clusters from 1980-2020 in a series of small
multiples. From these maps, we can pick out some general spatial
patterns over time. For example, we can see the expanse of neighborhoods
classified as majority White from 1980 significantly diminishes by 2020.
The share of Hispanics is shown to increase over time in the northern
sections of the City. Towards the East, we see neighborhoods generally
transition from predominantly White in 1980 to White and Mixed Race and
eventually to Asian by 2020. There are also some evident stable
clusters. For instance, the two clusters of predominantly black
neighborhoods in the South and Southeast appear quite table over time.
However, the cluster of majority Black neighborhoods in towards the
north of Manhattan is diminished, replaced by a Black and Hispanic
classification.

```{r smallmultiples,warning=FALSE, cache=TRUE, message=FALSE, fig.show='hold', fig.cap="Change in the cluster memmbership over five decades"}
decades <- c(1980, 1990, 2000, 2010, 2020)
  
# Filter and prepare data for mapping
tract_clusters_decades <- tract %>%
  left_join(census_long %>%
              select(TRTID10, year, label_6), by = "TRTID10") %>%
  filter(!is.na(label_6)) 

# Set factor levels for year to ensure proper ordering
tract_clusters_decades <- tract_clusters_decades %>%
  mutate(year = factor(year, levels = decades))

ggplot(tract_clusters_decades) +
  geom_sf(aes(fill = label_6), color = NA) +
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +
  facet_wrap(~ year, ncol = 5, strip.position = "bottom") +
theme_void() +
theme(
  plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
  strip.background = element_blank(),
  strip.text = element_text(size = 11, face = "italic", color = "gray20"),
  strip.placement = "outside",
  legend.position = "bottom",
  legend.direction = "horizontal",
  legend.title = element_blank(),
  legend.text = element_text(size = 9),
  legend.key.width = unit(1.2, "cm"),
  legend.key.height = unit(0.5, "cm")
) +
  scale_fill_manual(
    values = c(
      "3 White" = "#F0E442",
      "2 Hispanic" = "#D55E00",
      "1 Black" = "#0072B2",
      "5 Asian" = "#CC79A7",
      "4 Mixed" = "#009E73",
      "6 Black & Hispanic" = "#56B4E9"
    ),
    labels = c("Black", "Hispanic", "White", "Asian", "Mixed", "Black & Hispanic"),
    name = ""
  ) +
  labs(
    title = "Racial Cluster Maps by Decade (1980–2020)",
    x = "", y = ""
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 9),
    legend.key.width = unit(1.2, "cm"),
    legend.key.height = unit(0.5, "cm")
  ) +
  guides(
    fill = guide_legend(nrow = 1, byrow = TRUE, label.position = "bottom")
  )
```

### Sequence Analysis

Our objective with the sequence classification is to come up with a
typology of neighborhood sequences over time to describe general
pathways of change. Each neighborhood has a sequence of classes over
time, one of 6 categorical groups for each of the five decennial census
values from 1980-2020. The first step in the analysis is to convert the
data frame into a sequence for each neighborhood. We can see an
illustrative example of the longitudinal sequences from the first five
records. Classes are separated by a dash (-). There are 2122 sequences
in the dataset and 197 unique sequences; all sequences are displayed in
the plot below. Thus, the purpose of clustering the sequences is to
extract the general patterns present from the set of all sequences.


```{r createsequences, warning=FALSE, cache=TRUE, message=FALSE, fig.cap="Sequence of each neighborhood."}

# Convert the data from long to wide format to have a sequence for each neighborhood
# Assuming census_long is your data frame
census_wide <- tract_clusters_decades %>%
  st_drop_geometry() %>%
  select(TRTID10, year, label_6) %>%
  pivot_wider(names_from = year, values_from = label_6)

# Rename the columns to show only the year (remove "cluster_" prefix if it was added during previous steps)
colnames(census_wide) <- sub("cluster_", "", colnames(census_wide))

# Ensure columns are in the correct order by year
census_wide <- census_wide %>%
  select(TRTID10, `1980`, `1990`, `2000`, `2010`, `2020`)

# Ensure the sequence columns are factors
census_wide <- census_wide %>%
  mutate(across(starts_with("19") | starts_with("20"), as.factor))

# Check the distinct states (categories) in your sequences
unique_states <- unique(unlist(census_wide[, -1]))  # Exclude TRTID10 column
num_states <- length(unique_states)
print(unique_states)
print(num_states)  # Number of unique states

# Define the custom color palette with the correct colors
custom_palette <- c(
  "1 Black" = "#0072B2",
  "2 Hispanic" = "#D55E00",
  "3 White" = "#F0E442",
  "4 Mixed" = "#009E73",
  "5 Asian" = "#CC79A7",
  "6 Black and Hispanic" = "#56B4E9"
)


# Create the sequence object with the renamed columns
sequence_data <- seqdef(census_wide[, -1], cpal = custom_palette)  # Exclude the TRTID10 column

# Check the number of distinct sequences
num_sequences <- seqtab(sequence_data, idx = 0) %>% nrow
print(num_sequences)

# Display the first few sequences
head(sequence_data)

# Plot the sequences
seqIplot(sequence_data,                                             # Sequence object
         with.legend = "right",                                # Display legend on right side of plot
         cex.legend = 0.6,                                     # Change size of legend
         main = "Neighborhood Racial and Ethnic Trajectories") # Plot title

```



The next objective is to compute the dissimilarity between all
sequences. As described previously, we use the OMstrans algorithm to
preserve the ordering of events as we are most interested in describing
how neighborhoods have generally transitioned over time. We set a low
value (`0.1`) for the parameter `otto` in the `seqdist` command which is
the origin-transition trade-off weight. This emphasizes the ordering of
sequence states. We also set a high `indel` cost of `3`. Finally,
substitution costs are a function of the transition rate `TRATE` between
states. This places a lower cost on more frequent transitions.

The substitution cost matrix is shown in the Table below. The table indicates that more frequent transitions from, for example, Black & Hispanic to Black, Mixed to White, White to Mixed, and Black to Black and Hispanic have lower costs than rarer transitions from, for example, Asian to Black, Hispanic to Black, or Black to Asian. Generally, the more mixed race groups are more transitional white the majority race groups tend to transition through a mixed race state. This will become more evident as we examine the resulting sequence clusters. 

Once the cost matrix is established, we then cluster the sequences using
the dissimilarity matrix as an input to generate the typology. We follow
an iterative process in determining the optimal solution like the one
described above for the *k* means clustering. In short, multiple
solutions are tested and the resulting sequence clusters are visualized
to inspect for heterogeneity. Because our clustering and distance matrix
are optimized for describing transitions over time, we end up with one
cluster that contains all neighborhoods that remained constant over
time. These sequences are very different from one another in the
neighborhood types they describe, but they all represent a pathway or
sequences of no change.

To describe the resulting sequence clusters, we plot a Sequence
Frequency Plot for each cluster. We settled on 14 trajectory clusters
describing neighborhood racial and ethnic transitions from 1980 to 2020
in New York City. The Sequence Frequency Plots illustrate the sequences
belonging to each cluster and the sequence bars are scaled to visualize
the frequency of each sequence. A summary of the trajectories is as
follows:

1.  Hispanic Majority to Black & Hispanic
2.  Stability Cluster
3.  White & Mixed Race to Hispanic Majority
4.  White to White & Mixed Race to Hispanic Majority
5.  Black & Hispanic to Hispanic Majority
6.  White Majority to White & Mixed Race
7.  White & Mixed Race to Black & Hispanic
8.  Black & Hispanic to Black Majority
9.  White & Mixed Race to Asian Majority
10. White & Mixed Race to White Majority
11. Black & Hispanic to White & Mixed Race
12. White & Mixed Race to Asian Majority
13. Black Majority to Black & Hispanic
14. Black & Hispanic to Asian Majority

```{r clustersequences, warning=FALSE, cache=TRUE, message=FALSE}
# Pass this palette to seqdef
# Define sequence data with custom color palette
sequence_data <- seqdef(census_wide[, -1], cpal = custom_palette)  # Exclude the TRTID10 column

# Compute Optimal Matching (OM) distances using the TRATE cost method
costs <- seqcost(sequence_data, method = "TRATE")
om_distances <- seqdist(sequence_data, method = "OMstran", indel = 3, sm = costs$sm, otto = 0.1)

#Extract the substitution cost matrix
sub_matrix <- round(costs$sm, 2)  # round to 2 decimal places for clarity

# Convert to a data frame for export
sub_df <- as.data.frame(sub_matrix)

# Add row names as a column for better display
sub_df$From <- rownames(sub_df)
sub_df <- sub_df[, c("From", setdiff(names(sub_df), "From"))]

# Optional: reorder columns to match row order
sub_df <- sub_df[, c("From", rownames(sub_matrix))]

# Print to console
print(sub_df)


# Perform hierarchical clustering using Ward's method on the OM distances
clusterward <- agnes(om_distances, diss = TRUE, method = "ward")

# Define the number of clusters and assign clusters to the sequence data
num_clusters <- 14
clusters <- cutree(clusterward, k = num_clusters)
census_wide$sequence_cluster <- clusters



# Define all 14 cluster names
cluster_names_full <- c(
  "Hispanic Majority to Black & Hispanic",
  "Stability Cluster", 
  "White & Mixed Race to Hispanic Majority",
  "White to White & Mixed Race to Hispanic Majority",
  "Black & Hispanic to Hispanic Majority",
  "White Majority to White & Mixed Race",
  "White & Mixed Race to Black & Hispanic",
  "Black & Hispanic to Black Majority",
  "White & Mixed Race to Asian Majority",
  "White & Mixed Race to White Majority",
  "Black & Hispanic to White & Mixed Race",
  "White & Mixed Race to Asian Majority",
  "Black Majority to Black & Hispanic",
  "Black & Hispanic to Asian Majority"
)

# Modified function that creates separate plots for each cluster in the range
plot_sequence_clusters <- function(cluster_range) {
  for(i in cluster_range) {
    seqfplot(
      sequence_data[census_wide$sequence_cluster == i, ], 
      sortv = "from.start", 
      border = NA, 
      with.legend = "right",
      legend.prop = 0.2,
      legend.border = FALSE,
      main = paste("Cluster", i, ":", cluster_names_full[i]),
      cex.legend = 0.6  # Adjust this value: smaller = smaller text, larger = larger text
    )
  }
}

# Keep your existing calls - each will now create clean individual plots
plot_sequence_clusters(1:2)    # Creates 2 separate plots with individual titles
plot_sequence_clusters(3:4)    # Creates 2 separate plots with individual titles
plot_sequence_clusters(5:6)    # Creates 2 separate plots with individual titles
plot_sequence_clusters(7:8)    # Creates 2 separate plots with individual titles
plot_sequence_clusters(9:10)   # Creates 2 separate plots with individual titles
plot_sequence_clusters(11:12)  # Creates 2 separate plots with individual titles
plot_sequence_clusters(13:14)  # Creates 2 separate plots with individual titles
```

Of these 14 pathways, there are 3 that lead to the formation of a
neighborhood transitioning into a Hispanic Majority cluster by 2020.
This includes Cluster 3 - showing neighborhoods that went from being a
slight majority White, but with a mixture of other races in 1980 and
1990 to transitioning to majority Hispanic by 1990 or 2000. Some of
these sequences indicate a continued transition towards becoming
majority Asian in the later years. Further segmenting the sequences into
more clusters may have separated out those trajectories, but for the
sake of brevity, we leave them mixed in. Spatially, these are shown in
orange on the map below and can be see clustered in Staten Island,
Queens, and in the Northern portion of the city. They are also notably
adjacent to neighborhoods indicated by the red color, those representing
sequence cluster 4, transitioning from majority White to White and Mixed
Race and then to Majority Hispanic. This latter cluster might represent
the precursor to cluster 3, but are neighborhoods that made this
transition from majority White later, where the changes took time to
spatially spillover to adjacent neighborhoods, as indicated by the map.
Both of these pathways depict a transition from largely White
populations to largely Hispanic.

The third pathway depicting a transition to majority Hispanic is
distinct. It is represented by cluster 5 showing a transition from
either majority Black neighborhoods towards a mixed Hispanic and Black
group and eventually majority Hispanic or, beginning the 1980 time
stamp, in a more mixed Black and Hispanic state. Geographically, these
neighborhoods are shown more in the northern sections of Manhattan and
the Bronx.

From these two sets of sequence clusters, we can see that majority
Hispanic neighborhoods in New York City have emerged out of either
majority Black or Majority White neighborhoods over time.

```{r transition, fig.cap="Neighborhoods Following Pathway to Hispanic Majority with White County Borders.", echo=FALSE, warning=FALSE, message=FALSE}
# Define the Hispanic majority clusters for plotting
hispanic_majority_clusters <- census_wide %>%
  filter(sequence_cluster %in% c(4, 5, 6))  # Filter clusters 4, 5, and 6

# Join the filtered data with the tract shapefile based on TRTID10
tract_hispanic_majority <- tract %>%
  left_join(hispanic_majority_clusters, by = "TRTID10") %>% erase_water(area_threshold = 0.75)

# Add county borders and color to the plot
ggplot(tract_hispanic_majority) +
  geom_sf(aes(fill = factor(sequence_cluster)), color = NA) +  # Map sequence clusters
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +  # Add white county borders
  scale_fill_manual(values = c("#FB8072", "#80B1D3", "#FDB462"),  # Custom color palette
                    labels = c("3 White - White, Mixed Race - Hispanic", 
                               "5 Black - Black & Hispanic - Hispanic", 
                               "4 White, Mixed Race - Hispanic"), 
                    name = "Hispanic Majority Pathway") +
  labs(title = "", 
       x = "", y = "") +
  theme_void() +
  theme(legend.position = "right",  # Place legend on the right
        legend.direction = "vertical",  # Arrange legend items vertically
        legend.title = element_text(size = 12),  # Customize legend title size
        legend.key.width = unit(2, "cm"),  # Adjust legend key width
        legend.box = "vertical")  # Place legend title on top of the legend
```

There are also 3 pathways leading to an Asian majority neighborhood
type. These include clusters 9 and 12 which are also likely
continuations of longer trajectories, that show a gradual transition
from Majority White to White mixed race and eventually to Asian
Majority. Geographically, these are clustered in the northern section of
Queens. Sequence cluster 14 is more distinct in that the Asian majority
transitioned from the Black and Hispanic mixed group and spatially, they
are generally located in upper Manhattan and the Bronx.

```{r AsianPathways, fig.cap="Neighborhoods Following Pathway to Asian Majority", echo=FALSE, warning=FALSE, message=FALSE}
# Filter census data to include only clusters 14, 12, and 9 (Asian Majority Pathway clusters)
asian_majority_clusters <- census_wide %>%
  filter(sequence_cluster %in% c(14, 12, 9))  
  
# Join the filtered data with the tract shapefile (here, use TRTID10)
tract_asian_majority <- tract %>% erase_water(area_threshold = 0.75) %>% 
  left_join(asian_majority_clusters, by = "TRTID10")  # Join with spatial data

# Create a map of the neighborhoods with clusters 14, 12, and 9 with new labels
ggplot(tract_asian_majority) +
  geom_sf(aes(fill = factor(sequence_cluster)), color = NA) +  # Map sequence clusters
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +  # Add white county borders
  scale_fill_manual(values = c("#8DD3C7", "#FFFFB3", "#BEBADA"),  # Custom color palette for Asian Majority clusters
                    labels = c("14 White, Mixed Race to Black & Hispanic to Asian Majority", 
                               "12 White Majority to White Mixed Race to Asian", 
                               "9 White Mixed Race to Asian"), 
                    name = "Asian Majority Pathway") +
  labs(title = "", 
       x = "", y = "") +
  theme_void() +
  theme(legend.position = "right",  # Place legend at the bottom
        legend.direction = "vertical",  # Arrange legend items horizontally
        legend.title = element_text(size = 12),  # Customize legend title size
        legend.key.width = unit(2, "cm"),  # Adjust legend key width
        legend.box = "vertical")  # Place legend title on top of the legend
```

```{r whiteandblack}
# Filter data for Increasing White (clusters 10 and 11)
increasing_white_clusters <- census_wide %>% 
  filter(sequence_cluster %in% c(10, 11))  # Clusters 10 and 11 for increasing White

# Filter data for Increasing Black (clusters 1 and 8)
increasing_black_clusters <- census_wide %>%
  filter(sequence_cluster %in% c(1, 8))  # Clusters 1 and 8 for increasing Black

# Join the filtered data with the tract shapefile for each group
tract_increasing_white <- tract %>%  erase_water(area_threshold = 0.75) %>%
  left_join(increasing_white_clusters, by = "TRTID10")  # Join spatial data for increasing White

tract_increasing_black <- tract %>%   erase_water(area_threshold = 0.75) %>%
  left_join(increasing_black_clusters, by = "TRTID10")  # Join spatial data for increasing Black
```

There are two pathways for increasing both Black and White shares in a
neighborhoods. Neighborhoods that became increasingly White either
followed a trajectory from White mixed race to majority White (Cluster
10) or from either all Black or Black and Hispanic to White and Mixed
race (11). Notably, this transition largely took place within the past
1-2 decades, aligning with when gentrification trends became accentuated
in some cities, including New York. We see a clear cluster of this
latter group in Brooklyn, a borough whose gentrification trends have
been well documented [@chronopoulos2020s; @halasz2023between].

For the case of increasing Black populations, Cluster 1 shows a pathway
from Hispanic majority to mixed Black and Hispanic and Cluster 8 shows a
gradual transition from Black and Hispanic to majority Black; a trend
that largely begin towards the middle of the study period, around the
2000 census data mark. Spatially, the former is more dispersed, while
the latter is depicted more clearly in the southern neighborhoods of
Brooklyn.

```{r WhitePathways, fig.cap="Neighborhoods Following Pathway to White Majority", echo=FALSE, warning=FALSE, message=FALSE}
# Adjust plotting settings to improve map visibility
# Create the map for Increasing White, zoomed in on the five boroughs
ggplot(tract_increasing_white) +
  geom_sf(aes(fill = factor(sequence_cluster)), color = NA) +  # Map sequence clusters
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +  # Add white county borders
  scale_fill_manual(values = c("#8DD3C7", "#FFFFB3"),  # Custom color palette for White clusters
                    labels = c("White Mixed Race to Majority White", 
                               "Black and Hispanic to White Mixed Race"), 
                    name = "Increasing White Pathway") +
  coord_sf(xlim = c(-74.3, -73.7), ylim = c(40.5, 40.9), expand = FALSE) +  # Zoom into the NYC area
  theme_void() +
  theme(legend.position = "right",  # Place legend at the right
        legend.direction = "vertical",  # Arrange legend items vertically
        legend.title = element_text(size = 12),  # Customize legend title size
        legend.key.width = unit(2, "cm"),
        legend.box = "vertical")  # Place legend title on top
```

```{r Black_Pathways, fig.cap="Neighborhoods Following Pathway to Black Majority", echo=FALSE, warning=FALSE, message=FALSE}
# Create the map for Increasing Black, zoomed in on the five boroughs
ggplot(tract_increasing_black) +
  geom_sf(aes(fill = factor(sequence_cluster)), color = NA) +  # Map sequence clusters
  geom_sf(data = nyc_counties, fill = NA, color = "black", size = 0.5) +  # Add white county borders
  scale_fill_manual(values = c("#FB8072", "#80B1D3"),  # Custom color palette for Black clusters
                    labels = c("Hispanic Majority to Black & Hispanic", 
                               "Black and Hispanic to Black Majority"), 
                    name = "Increasing Black Pathway") +
  coord_sf(xlim = c(-74.3, -73.7), ylim = c(40.5, 40.9), expand = FALSE) +  # Zoom into the NYC area
  theme_void() +
  theme(legend.position = "right",  # Place legend at the right
        legend.direction = "vertical",  # Arrange legend items vertically
        legend.title = element_text(size = 12),  # Customize legend title size
        legend.key.width = unit(2, "cm"),
        legend.box = "vertical")  # Place legend title on top

```

Finally, of the sequences depicting no change from 1980 to 2020, shown
in Cluster 2, the majority of those are for tracts with a racial
majority: White, Hispanic, and Black. But those are followed by two
stable sequences of neighborhoods with some racial diversity. The first
is the majority White, but with a mixture of other races and the second
is the mixed Black and Hispanic cluster. There is some debate in the
literature whether racially mixed neighborhoods can be stable over time,
or are they simply depicting a point along a pathway of change when one
group will eventually become a majority. Research has shown that highly
diverse neighborhoods are quite unstable over time, likely to transition
to a less diverse state. In particular the transition from predominantly
White towards majority Hispanic, results in a period of unstable racial
mixture while that transition takes place [@wright2020instability].
Others have identified a small, but persistent set of racially diverse
neighborhoods throughout the United States, but particularly in racially
diverse metropolitan areas [@hipp2023persistent]. Here, we find some
evidence of stable, non-racially homogeneous Census Tracts.

## Conclusions {#sec-conclusion}

This analysis showcases a method for developing and visualizing a
typology of neighborhood change pathways. We used a case study of
decennial racial and ethnic changes in New York City census tracts from
1980-2020. The workflow first involves developing a cross-sectional
typology of classes describing the racial mixture of neighborhoods. To
do so, we used an unsupervised classification approach, *k*-means to
derive six such clusters.

We demonstrated how determining the number of clusters often falls to
more of an art than a precise science as our final clustering for this
analysis exceeded the mathematically optimal three clusters, but
provided us with more nuance on the racial mixture of neighborhoods. We
performed the cluster analysis on all census tracts in the city for each
of the five decennial census time stamps at once to ensure a temporally
consistent set of groupings. We then created sequences of neighborhood
clusters over the study period and developed a typology of sequences
that grouped them based on the similarity of how they changed over time.
To do so, we used the OMtrans algorithm for computing sequence
dissimilarity to ensure that the ordering or sequencing of events was
preserved. Finally, we mapped sequence clusters to spatially visualize
neighborhoods that followed similar pathways of change. For our case
study of the largest city in the United States, and one of the most
diverse, we observed a decline in the number of majority White census
tracts over time. We identified several pathways of change leading to
majority Hispanic neighborhoods - emerging either out of previously
majority White or Black neighborhoods. We also observed pathways leading
to majority Asian tracts, transitioning from Black and Hispanic
neighborhoods, largely in Queens. A smaller share of neighborhoods
became either increasingly White or Black. Neighborhoods that saw an
increase in the share of Whites, saw a notable increase in the
transition from Black and Hispanic to White and mixed race over the past
two decades. Neighborhoods that increased in the share of Blacks largely
transitioned from Black and Hispanic to majority Black or from White and
mixed race to Black and Hispanic. 

The strength of the sequence analysis technique lies in its ability to clearly visualize common pathways of neighborhood change. One of its limitations, however, is the need to segment continuous, longitudinal data into discrete, categorical states. In this article, we devoted attention to this step, illustrating the use of a *k*-means algorithm as an unsupervised classification method for grouping multiple variables. This phase involves decisions, particularly, the number of classes, as that directly influences the resulting sequences and subsequent interpretations. Alternative methods for clustering time series data that preserve the continuous nature of the data may be preferable in cases where the number of neighborhood variables is limited [@delmelle2025time].

Beyond describing trajectories, sequence analysis also opens up avenues for further inquiry. The resulting sequence clusters can serve as the basis for additional analyses like modeling the predictors of specific trajectories or analyzing their spatial patterns using categorical spatial autocorrelation measures such as the join-count statistic.